---
title: "2023"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options: 
  chunk_output_type: inline
---

```{r include=FALSE}
library(tidyverse)
library(plotly)
library(ggplot2)
library(gt)
```

```{r include=FALSE}
spotify2023_df = 
  read_csv("spotify-2023.csv") |> 
  janitor::clean_names() |> 
  rename_with(~ gsub("_percent", "", .), contains("_percent")) |> 
  rename(artist = artist_s_name)
view(spotify2023_df)
```
# Exploratory Data Analysis
### Discussion of Raw Data

The spotify-2023.csv dataset contains information about the most streamed songs on Spotify in 2023. Each observation includes basic information such as the track name, artist name, release date, and number of streams, as well as additional characteristics like bpm, key, mode, and other audio features. Popularity is measured by presence in Spotify playlists and charts, and across playlists and charts of other streaming platforms including Apple, Deezer, and Shazam.

```{r echo = TRUE, results = 'hide', include=FALSE}
spotify2023_df = 
  read_csv("data/spotify-2023.csv") |> 
  janitor::clean_names() |> 
  rename_with(~ gsub("_percent", "", .), contains("_percent")) |> 
  rename(artist = artist_s_name) |> 
  mutate(streams = ifelse(track_name == "Love Grows (Where My Rosemary Goes)", NA, streams),
         streams = as.numeric(streams))

view(spotify2023_df)
```

The dataset was already quite clean, but we did come across a few minor issues. For example, the track "Love Grows" by Edison Lighthouse contained a "streams" observation that was not a numerical value but rather a very long character value and was preventing the aggregation of streaming data. To address this, the value was set as "NA".

The dataset includes the following variables:

- `track_name`: Name of the song.
- `artist`: Name of the artist(s).
- `artist_count`: Number of artists contributing to the song.
- `released_year`: Year the song was released.
- `released_month`: Month the song was released.
- `released_day`: Day of the month the song was released.
- `in_spotify_playlists`: Number of Spotify playlists the song is included in.
- `in_spotify_charts`: Presence and rank of the song on Spotify charts.
- `streams`: Total number of streams on Spotify.
- `in_apple_playlists`: Number of Apple playlists the song is included in.
- `in_apple_charts`: Presence and rank of the song on Apple charts.
- `in_deezer_playlists`: Number of Deezer playlists the song is included in.
- `in_deezer_charts`: Presence and rank of the song on Deezer charts.
- `in_shazam_charts`: Presence and rank of the song on Shazam charts.
- `bpm`: Beats per minute, a measure of song tempo.
- `key`: Key of the song.
- `mode`: Mode of the song (major or minor).
- `danceability`: Percentage indicating how suitable the song is for dancing.
- `valence`: Positivity of the song's musical content.
- `energy`: Perceived energy level of the song.
- `acousticness`: Amount of acoustic sound in the song.
- `instrumentalness`: Amount of instrumental content in the song.
- `liveness`: Presence of live performance elements.
- `speechiness`: Amount of spoken words in the song.

### Summary Data

*Summary Statistics*
```{r}
summary_table = spotify2023_df |> 
  summarize(
    spotify_streams_mean = mean(streams, na.rm = TRUE),
    spotify_streams_median = median(streams, na.rm = TRUE),
    spotify_streams_iqr = IQR(streams, na.rm = TRUE),
    
    spotify_playlists_mean = mean(in_spotify_playlists, na.rm = TRUE),
    spotify_playlists_median = median(in_spotify_playlists, na.rm = TRUE),
    spotify_playlists_iqr = IQR(in_spotify_playlists, na.rm = TRUE),
    
    apple_playlists_mean = mean(in_apple_playlists, na.rm = TRUE),
    apple_playlists_median = median(in_apple_playlists, na.rm = TRUE),
    apple_playlists_iqr = IQR(in_apple_playlists, na.rm = TRUE),
    
    deezer_playlists_mean = mean(in_deezer_playlists, na.rm = TRUE),
    deezer_playlists_median = median(in_deezer_playlists, na.rm = TRUE),
    deezer_playlists_iqr = IQR(in_deezer_playlists, na.rm = TRUE)
  ) |> 
  pivot_longer(
    cols = everything(),
    names_to = c("Platform", "Statistic"),  
    names_pattern = "(.*)_(.*)"
  ) |> 
  mutate(
    Platform = case_when(
      Platform == "spotify_streams" ~ "Spotify Streams",
      Platform == "spotify_playlists" ~ "# of Spotify Playlists",
      Platform == "apple_playlists" ~ "# of Apple Playlists",
      Platform == "deezer_playlists" ~ "# of Deezer Playlists",
      TRUE ~ Platform)
  ) |> 
  pivot_wider(
    names_from = Statistic,               
    values_from = value
  )

summary_table |> 
  gt() |> 
  tab_header(
    title = "Summary Statistics Across Platforms",
    subtitle = "Mean, Median, and Interquartile Range"
  ) |> 
  cols_label(
    Platform = "Platform",
    mean = "Mean",
    median = "Median",
    iqr = "Interquartile Range"
  ) |> 
  fmt_number(
    columns = c(mean, median, iqr),
    decimals = 0
  ) |> 
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels()
  )
```

*Interpretation*
Total number of streams was only available for Spotify and displays a much greater mean than median, indicating a right-skewed distribution. The mean, median, and IQR for number of playlists each song was on are significantly larger for Spotify than Apple and Deezer. This suggests that the total number of playlists on Spotify is much larger than the total number of playlists on Apple or Deezer. The distributions of number of playlists for Spotify, Apple, and Deezer are also right-skewed.

*Top 10 Streamed Songs*
```{r top 10 streams}
top10streamed = spotify2023_df |>
  arrange(desc(streams)) |> 
  slice_head(n = 10)

table_top10streamed = top10streamed |>
  select(track_name, artist, streams) |>
         gt() |>
         tab_header(
           title = "Top 10 Streamed Songs on Spotify in 2024") |>
         fmt_number(
           columns = streams,
           decimals = 0
         ) |>
         cols_label(
           track_name = "Song Title",
           artist = "Artist",
           streams = "Streams"
         )

table_top10streamed
```

*Top 10 Songs by Playlist Count*
```{r top 10 playlist}
top10playlist <- spotify2023_df |>
  mutate(
    cumulative_playlist = rowSums(across(c(in_spotify_playlists, in_apple_playlists, in_deezer_playlists), ~ replace_na(.x, 0)))
  ) |> 
  arrange(desc(cumulative_playlist)) |>
  slice_head(n = 10)

table_top10playlist <- top10playlist |>
  select(track_name, artist, cumulative_playlist) |>
  gt() |>
  tab_header(
    title = "Top 10 Songs By Presence on Playlists",
    subtitle = "Aggregated across Spotify, Apple Music, and Deezer"
  ) |>
  fmt_number(
    columns = cumulative_playlist,
    decimals = 0
  ) |>
  cols_label(
    track_name = "Song Title",
    artist = "Artist",
    cumulative_playlist = "Cumulative Playlist Count"
  )

table_top10playlist
```

*Top 10 Songs by Rank Across Platforms*
```{r top 10 rank}
top10charts <- spotify2023_df |>
  mutate(
    chart_presence_score = rowSums(across(c(in_spotify_charts, in_apple_charts, in_deezer_charts, in_shazam_charts), ~ replace_na(.x, 0)))
  ) |>
  arrange(desc(chart_presence_score)) |>
  slice_head(n = 10)

table_top10charts <- top10charts |>
  select(track_name, artist, chart_presence_score) |>
  gt() |>
  tab_header(
    title = "Top 10 Songs By Chart Presence",
    subtitle = "Aggregated across Spotify, Apple Music, Deezer, and Shazam"
  ) |>
  fmt_number(
    columns = chart_presence_score,
    decimals = 0
  ) |>
  cols_label(
    track_name = "Song Title",
    artist = "Artist",
    chart_presence_score = "Cumulative Chart Presence"
  )

table_top10charts
```

The top 10 songs vary among top streamed, presence on playlists, and presence on charts. There are a few consistencies across Top 10 Streamed Songs and Top 10 Songs By Presence on Playlists, such as "Blinding Lights" and "One Dance", but overall the top songs differ completely by these measures.

*Top 10 Artists by Total Number of Spotify Streams*
```{r top 10 artist}
top10artists <- spotify2023_df |>
  group_by(artist) |>
  summarise(total_streams = sum(streams, na.rm = TRUE)) |>
  arrange(desc(total_streams)) |>
  slice_head(n = 10)

table_top10artists <- top10artists |>
  gt() |>
  tab_header(
    title = "Top 10 Artists By Total Spotify Streams",
    subtitle = "Based on aggregated stream counts"
  ) |>
  fmt_number(
    columns = total_streams,
    decimals = 0
  ) |>
  cols_label(
    artist = "Artist",
    total_streams = "Total Streams"
  )

table_top10artists
```

*Top 3 Songs by Playlist Count Between Platforms*
```{r}
top3_combined <- bind_rows(
  spotify2023_df |>
    arrange(desc(in_spotify_playlists)) |>
    slice_head(n = 3) |>
    mutate(platform = "Spotify", playlist_count = in_spotify_playlists),
  
  spotify2023_df |>
    arrange(desc(in_apple_playlists)) |>
    slice_head(n = 3) |>
    mutate(platform = "Apple Music", playlist_count = in_apple_playlists),
  
  spotify2023_df |>
    arrange(desc(in_deezer_playlists)) |>
    slice_head(n = 3) |>
    mutate(platform = "Deezer", playlist_count = in_deezer_playlists)
)

table_top3_combined <- top3_combined |>
  select(platform, track_name, artist, playlist_count) |>
  gt() |>
  tab_header(
    title = "Top 3 Songs by Playlist Count Across Platforms",
    subtitle = "Spotify, Apple Music, and Deezer"
  ) |>
  fmt_number(
    columns = playlist_count,
    decimals = 0
  ) |>
  cols_label(
    platform = "Platform",
    track_name = "Song Title",
    artist = "Artist",
    playlist_count = "Playlist Count"
  ) |>
  tab_spanner(
    label = "Playlist Data",
    columns = c(track_name, artist, playlist_count)
  ) |>
  opt_row_striping()

table_top3_combined
```
Assessing popularity through number of playlists a song is included on showed minimal overlap between Spotify, Apple Music, and Deezer. Only top 3 were included for the sake of simplicity in the table. "Get Lucky - Radio Edit" by Pharrell Williams, Nile Rodgers, and Daft Punk is among the top 3 songs by playlist count on both Spotify and Deezer.

### Visualizations
#### A Year in Sound: What Made 2023's Music Stand Out?


**Tuning Into Music: A Journey Through Keys and Modes**

*Distribution of Spotify Streams by Key*
```{r}
streams_by_key <- spotify2023_df |>
  filter(!is.na(key)) |>
  group_by(key) |>
  summarise(total_streams = sum(streams, na.rm = TRUE)) |>
  arrange(desc(total_streams))

streams_by_key_plot <- ggplot(streams_by_key, aes(x = reorder(key, total_streams), y = total_streams, fill = total_streams)) +
  geom_bar(stat = "identity", color = "pink", alpha = 0.7) +   
  scale_y_continuous(labels = scales::comma) +                  
  scale_fill_gradient(low = "lightgreen", high = "darkgreen") + 
  labs(
    title = "Total Streams by Musical Key",
    x = "Musical Key",
    y = "Total Streams",
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "none"
  )

streams_by_key_plot
```

The most common key for top songs is C#, whereas the least popular is D#. The distribution among other keys of music appears to be relatively even.

*Counts of Songs by Musical Key*
```{r results = 'hide'}
key_counts <- as.data.frame(table(spotify2023_df$key))

str(key_counts)

colnames(key_counts) <- c("key", "count")

p <- ggplot(key_counts, aes(x = key, y = count, 
                            fill = as.numeric(key),  
                            text = paste("Key: ", key, 
                                         "<br>Count: ", count))) +
  geom_bar(stat = "identity", color = "pink", width = 0.7) +  
  labs(title = "Count of Songs by Musical Key",
       x = "Musical Key",
       y = "Count of Songs") +
  theme_minimal() +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen", name = "Key") +  
  theme(legend.position = "none")  # Remove the legend

interactive_plot <- ggplotly(p, tooltip = "text")

interactive_plot
```

The musical key with the highest song count is C#(120), with G#(96) and G(91) being the next highest. The lowest count key was D# with 33 songs. There was slight variance between other keys, however most were comparable. 

*Distribution of Spotify Streams by Mode*
```{r}
streams_by_mode <- spotify2023_df |>
  filter(!is.na(mode)) |>
  group_by(mode) |>
  summarise(total_streams = sum(streams, na.rm = TRUE)) |>
  arrange(desc(total_streams))

streams_by_mode_plot <- ggplot(streams_by_mode, aes(x = reorder(mode, total_streams), y = total_streams, fill = total_streams)) +
  geom_bar(stat = "identity", color = "pink", alpha = 0.7) + 
  scale_y_continuous(labels = scales::comma) +               
  scale_fill_gradient(low = "lightgreen", high = "darkgreen") + 
  labs(
    title = "Total Streams by Musical Mode",
    x = "Musical Mode",
    y = "Total Streams"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "none"
  )

streams_by_mode_plot
```

More streams occurred among songs in a major key, suggesting that upbeat, positive, and joyful songs were more popular than darker, melancholic, and emotional songs. 


**Tempo Trends: Fast or Slow Beats?** 

*Distribution of Spotify Streams by BPM*
```{r}
streams_by_bpm <- spotify2023_df |>
  filter(!is.na(bpm)) |>
  group_by(bpm) |>
  summarise(total_streams = sum(streams, na.rm = TRUE)) |>
  arrange(bpm) 

streams_by_bpm_plot <- ggplot(streams_by_bpm, aes(x = bpm, y = total_streams, fill = total_streams)) +
  geom_bar(stat = "identity", color = "pink", alpha = 0.7) +   
  scale_y_continuous(labels = scales::comma) +                  
  scale_fill_gradient(low = "lightgreen", high = "darkgreen") +  
  scale_x_continuous(breaks = seq(floor(min(streams_by_bpm$bpm)/10)*10, ceiling(max(streams_by_bpm$bpm)/10)*10, by = 10)) +  
  labs(
    title = "Total Streams by BPM",
    x = "Beats per Minute (BPM)",
    y = "Total Streams"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    panel.grid.minor = element_blank()  
  )

streams_by_bpm_plot
```

The distribution of total streams by BPM shows that most top songs have a BPM somewhere between 90-150. The data show a slightly right-skewed distribution, suggesting that songs with higher BPMs may be less popular. The sweet spot appears to be 120 BPM, where the greatest number of streams were recorded in 2023.

*Danceability vs. Energy*
```{r}
spotify2023_df$artist <- iconv(spotify2023_df$artist, to = "ASCII", sub = "")
spotify2023_df$track_name <- iconv(spotify2023_df$track_name, to = "ASCII", sub = "")

p <- ggplot(spotify2023_df, aes(x = danceability, y = energy, color = danceability,
                      text = paste("Track: ", track_name, 
                                   "<br>Artist: ", artist, 
                                   "<br>Danceability: ", round(danceability, 2),
                                   "<br>Energy: ", round(energy, 2)))) +
  geom_point(alpha = 0.7) +
  scale_color_gradient(low = "lightgreen", high = "darkgreen") +
  labs(title = "Groove Meter: Danceability vs. Energy",
       x = "Danceability",
       y = "Energy") +
  theme_minimal()

interactive_plot <- ggplotly(p, tooltip = "text")

interactive_plot
```

This plot shows where songs stand with danceability and energy ratings. The main cluster of songs occurs between the 70-90 range on danceability and  50-90 range on energy. Few songs rank both low in energy and danceability. 


*Energy vs Danceability*
```{r}
p <- ggplot(spotify2023_df) +
  geom_density(aes(x = energy, fill = "energy", color = "pink"), alpha = 0.5) +  
  geom_density(aes(x = danceability, fill = "danceability", color = "pink"), alpha = 0.5) +  
  scale_fill_manual(values = c("energy" = "darkgreen", "danceability" = "lightgreen")) +  # Fill colors
  scale_color_manual(values = c("pink" = "pink")) + 
  labs(title = "Fueling the Beat: The Energy-Danceability Connection",
       x = "Value",
       y = "Density") +
  theme_minimal() +
  guides(fill = guide_legend(title = "Dynamics")) +  
  guides(color = "none")  

print(p)
```

This plot further explains the overlap of danceability and energy through an overlapped plot. 


**Multidimensional World of Sound**

*Liveness vs Speechiness* 
```{r}
scatter_plot_liveness_speechiness <- ggplot(
  spotify2023_df, 
  aes(
    x = liveness, 
    y = speechiness, 
    text = paste(
      "Track:", track_name, 
      "<br>Artist:", artist, 
      "<br>Liveness:", round(liveness, 2), 
      "<br>Speechiness:", round(speechiness, 2)
    ),
    color = liveness  
  )
) +
  geom_point(alpha = 0.7, size = 3) +
  scale_color_gradient(low = "lightgreen", high = "darkgreen") +  
  scale_x_continuous(limits = c(0, 100)) +  
  scale_y_continuous(limits = c(0, 75)) +  
  labs(
    title = "From Silence to Sound: Liveness and Speechiness",
    x = "Liveness",
    y = "Speechiness"
  ) +
  theme_minimal() +
  theme(legend.position = "none")  

interactive_scatter_liveness_speechiness <- ggplotly(scatter_plot_liveness_speechiness, tooltip = "text")
interactive_scatter_liveness_speechiness
```
This plot shows a dense cluster in low liveness and speechiness, indicating that listeners likely prefer music that is not live and with fewer words. 


*Vocal vs. Instrumental*
```{r}
track_counts_df <- as.data.frame(table(ifelse(spotify2023_df$instrumentalness > 0.5, "Instrumental", "Vocal")))

colnames(track_counts_df) <- c("Status", "Count")


interactive_pie_chart <- plot_ly(track_counts_df, labels = ~Status, values = ~Count, type = 'pie',
                                  textinfo = 'label+percent', # This shows both label and percentage
                                  marker = list(colors = c('darkgreen', 'lightgreen'))) %>%
  layout(
    title = "The Sound Spectrum: Vocal or Instrumental?",
    showlegend = F
  )

interactive_pie_chart
```
This pie chart breaks down the percent of vocal(90.9%) to instrumental(9.13%) tracks, showing a clear dominance to those that are vocal. 


**Data Wrapped: The Ultimate Chart-Toppers of 2023**

*Top 10 Artists with the Most Danceable Songs (Danceability > 50)*
```{r}
top_artists_danceability <- spotify2023_df %>%
  filter(danceability > 50) %>%  
  group_by(artist) %>%
  summarise(danceable_songs = n()) %>%
  arrange(desc(danceable_songs)) %>%
  slice_head(n = 10) 

ggplot(top_artists_danceability, aes(x = reorder(artist, danceable_songs), y = danceable_songs, fill = danceable_songs)) +
  geom_bar(stat = "identity") +
  coord_flip() +  
  labs(
    title = "Top 10 Artists with the Most Danceable Songs (Danceability > 50)",
    x = "Artist",
    y = "Number of Danceable Songs"
  ) +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen") +  # Green gradient added here
  theme_minimal() +
  theme(legend.position = "none")
```

*Top 10 Artists with the Most Songs by Count*
```{r}
top_artists_songs <- spotify2023_df %>%
  group_by(artist) %>%
  summarise(songs_count = n()) %>%
  arrange(desc(songs_count)) %>%
  slice_head(n = 10)

ggplot(top_artists_songs, aes(x = reorder(artist, songs_count), y = songs_count, fill = songs_count)) +
  geom_bar(stat = "identity") +
  coord_flip() +  
  labs(
    title = "Top 10 Artists with the Most Songs in 2023",
    x = "Artist",
    y = "Number of Songs"
  ) +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen") +  # Green gradient
  theme_minimal() +
  theme(legend.position = "none")

```

*Top 10 Tracks with the Most Appearances in Spotify Charts*
```{r}
top_charts <- spotify2023_df %>%
  filter(in_spotify_charts > 0) %>%  
  arrange(desc(in_spotify_charts)) %>%  
  slice_head(n = 10)  

ggplot(top_charts, aes(x = reorder(track_name, in_spotify_charts), y = in_spotify_charts, fill = in_spotify_charts)) +
  geom_bar(stat = "identity") +
  coord_flip() +  
  labs(
    title = "Top 10 Tracks with the Most Appearances in Spotify Charts",
    x = "Song",
    y = "Number of Appearances in Charts"
  ) +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen") +  
  theme_minimal() +
  theme(legend.position = "none")  
```

*Top 10 Tracks with the Highest Energy*
```{r}
top_energetic_songs <- spotify2023_df %>%
  arrange(desc(energy)) %>%  
  slice_head(n = 10)  

ggplot(top_energetic_songs, aes(x = reorder(track_name, energy), y = energy, fill = energy)) +
  geom_bar(stat = "identity") +
  coord_flip() +  
  labs(
    title = "Top 10 Most Energetic Songs",
    x = "Song",
    y = "Energy"
  ) +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen") +  
  theme_minimal() +
  theme(legend.position = "none")
```


*Top 10 Songs in the Most Playlists*
```{r}
top_songs_playlists <- spotify2023_df %>%
  filter(in_spotify_playlists > 0) %>%  
  group_by(track_name) %>%  
  summarise(
    playlists_count = sum(in_spotify_playlists)  
  ) %>%
  arrange(desc(playlists_count)) %>%  
  slice_head(n = 10)  

ggplot(top_songs_playlists, aes(x = reorder(track_name, playlists_count), y = playlists_count, fill = playlists_count)) +
  geom_bar(stat = "identity") +
  coord_flip() +  
  labs(
    title = "Top 10 Songs in the Most Playlists",
    x = "Song",
    y = "Total Playlist Appearances"
  ) +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen") +  
  theme_minimal() +
  theme(legend.position = "none")  

```

These top ten charts display popular songs and artists that made the charts, various Spotify playlists, and who created iconic, danceworthy music!

