---
title: "2023 and 2024"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide 
editor_options: 
  chunk_output_type: console
---
```{r include=FALSE}
library(tidyverse)
library(plotly)
library(ggplot2)
library(gt)
library(scales)
library(broom)
library(knitr)
library(kableExtra)
```
# Exploratory Data Analysis
### Discussion of Raw Data
This page will focus on data that has merged both the 2023 and 2024 Spotify datasets. While the 2023 dataset has more qualitative data about the tracks themselves (e.g., bpm, key, danceability, valence, etc.), the 2024 dataset has a lot more information about the popularity of the tracks across several platforms (e.g. TikTok, etc.). We are specifically interested in tracks that are included in both of these datasets. The 2023 dataset has 953 tracks included, while the 2024 dataset has 4425 tracks. For more information about the individual datasets, visit [2023](2023.html) and [2024](2024.html).

```{r echo=TRUE, include=FALSE}
spotify2024_df= read_csv("data/2024_data.csv") |>
                janitor::clean_names() |>
                select(-tidal_popularity)|>
  mutate(
    track = iconv(track, from = "UTF-8", to = "ASCII//TRANSLIT"),
    artist = iconv(artist, from = "UTF-8", to = "ASCII//TRANSLIT")
  ) |> 
  separate(
    col = release_date,
    into = c("released_month", "released_day", "released_year"),
    sep = "/",
    convert = TRUE  
  ) |> 
    rename(
    `2024_spotify_streams` = spotify_streams,
    `2024_spotify_playlist_count` = spotify_playlist_count,
    `2024_apple_music_playlist_count` = apple_music_playlist_count,
    `2024_deezer_playlist_count` = deezer_playlist_count,
    `2024_tik_tok_posts`= tik_tok_posts,
     `2024_tik_tok_likes`= tik_tok_likes,
     `2024_tik_tok_views`= tik_tok_views,
  )|> 
  filter(artist!= "xSyborg") |> 
  select(track, released_month, released_day, released_year, `2024_spotify_streams`, `2024_spotify_playlist_count`, `2024_apple_music_playlist_count`,
    `2024_deezer_playlist_count`, `2024_tik_tok_posts`, `2024_tik_tok_likes`, `2024_tik_tok_views`)

spotify2023_df = 
  read_csv("data/spotify-2023.csv") |> 
  janitor::clean_names() |> 
  rename_with(~ gsub("_percent", "", .), contains("_percent")) |> 
  rename(artist = artist_s_name) |> 
  mutate(streams = ifelse(track_name == "Love Grows (Where My Rosemary Goes)", NA, streams),
         streams = as.numeric(streams)) |> 
  rename(
    track=track_name,
    '2023_spotify_streams' = streams,
    '2023_spotify_playlist_count' = in_spotify_playlists,
    '2023_apple_music_playlist_count' = in_apple_playlists,
    '2023_deezer_playlist_count' = in_deezer_playlists
    ) |> 
  select(track, artist, released_month, released_day, released_year, '2023_spotify_streams', '2023_spotify_playlist_count', '2023_apple_music_playlist_count', '2023_deezer_playlist_count', bpm, key, mode, danceability, valence, energy, acousticness, liveness, speechiness )
```


```{r}
# Merging datasets
spotify_df=
  inner_join(spotify2023_df, spotify2024_df, by= c("track", "released_month", "released_day", "released_year"), relationship="many-to-many")
```

```{r echo=TRUE, include=FALSE}
# Info about duplicate tracks in merged file
spotify_df |> 
  group_by(track) |> 
filter(n() > 1) |> 
  arrange(track)
```

```{r}
# Can also use the dataset in a longer format.
spotify_long_df=
  spotify_df |>
  select(
     `2023_spotify_streams`,  `2023_spotify_playlist_count`, `2023_apple_music_playlist_count`, `2023_deezer_playlist_count`, `2024_spotify_streams`, `2024_spotify_playlist_count`,    
`2024_apple_music_playlist_count`, `2024_deezer_playlist_count`, everything()     
  ) |> 
    pivot_longer(
    cols = c(`2023_spotify_streams`,  `2023_spotify_playlist_count`, `2023_apple_music_playlist_count`, `2023_deezer_playlist_count`, `2024_spotify_streams`, `2024_spotify_playlist_count`,    
`2024_apple_music_playlist_count`, `2024_deezer_playlist_count`, `2024_tik_tok_posts`,             `2024_tik_tok_likes`, `2024_tik_tok_views`),
    names_to = c("year", "metric"),                 
    names_pattern = "(\\d{4})_(.*)",    
   values_to = "value"  
  )
```

### Data Cleaning and Merging
Prior to merging the datasets, in the 2023 dataset, we renamed  `track_name` to `track`, `streams` to `2023_spotify_streams`, `in_spotify_playlists` to `2023_spotify_playlist_count`, `in_apple_playlists` to `2023_apple_music_playlist_count`, and`in_deezer_playlists` to `2023_deezer_playlist_count`. In the 2024 dataset, we separated `release_date` into `released_year`, `released_month`, `released_day`, changed `spotify_streams` to `2024_spotify_streams`, `spotify_playlist_count`  to `2024_spotify_playlist_count`, `apple_music_playlist_count` to `2024_apple_music_playlist_count`, and `deezer_playlist_count` to `2024_deezer_playlist_count`. Variables that were removed from the 2023 dataset prior to merging include the following: `artist_count`, `in_spotify_charts`, `in_apple_charts`, `in_deezer_charts`, `in_shazam_charts`, and `instrumentalness`. Variables that were removed from the 2024 dataset prior to merging include the following: `album_name`, `isrc`, `all_time_rank`, `track_score`, `spotify_playlist_reach`, `spotify_popularity`, `you_tube_views`, `you_tube_likes`, `air_play_spins`, `sirius_xm_spins`, `deezer_playlist_reach`, `amazon_playlist_count`, `pandora_streams`, and `pandora_track_stations`. 

Datasets were merged based on `track`, `released_year`, `released_month`, and `released_day`. The `artist` variable for the 2023 was kept because the 2023 dataset includes the artists that are featured on the track as well. 

When merging the datasets (through inner_join), we found that there were 17 duplicate songs where the duplicates all contain information for both 2023 and 2024 Spotify streams, Spotify playlist count, and Apple Music playlist count. Some duplicates, however, have varying missing information about 2024 Deezer playlist count and/or TikTok data. Some other notable discrepancies we found among the 17 tracks where the following:
- “Daylight” and “Numb” are duplicate track names but are from different artists 
- “My Universe” has limited information in both duplicates
- “SNAP” has two duplicates (4 total) 
- “SPIT IN MY FACE” is a duplicate but they are in different keys/have different song details (e.g., danceability, valence, etc.)


The merged 2023/2024 dataset (spotify_df) includes the following 25 variables:

  - `track`: Name of the song.
  - `artist`: Name of the artist(s) of the song.
  - `released_year`: Year the song was released.
  - `released_month`: Month the song was released.
  - `released_day`: Day of the month the song was released.
  - `2024_spotify_streams`: Total number of streams on Spotify in 2024.
  - `2024_spotify_playlist_count`: Number of Spotify playlists the song is included in in 2024.
  - `2024_apple_music_playlist_count`: Number of Apple Music playlists the song is included in in 2024.
  - `2024_deezer_playlist_count`: Number of Deezer playlists the song is included in in 2024.
  - `2023_spotify_streams`: Total number of streams on Spotify in 2023.
  - `2023_spotify_playlist_count`: Number of Spotify playlists the song is included in in 2023.
  - `2023_apple_music_playlist_count`: Number of Apple Music playlists the song is included in in 2023.
  - `2023_deezer_playlist_count`: Number of Deezer playlists the song is included in in 2023.
  - `2024_tik_tok_posts`: Number of TikTok posts featuring the song in 2024.
  - `2024_tik_tok_likes`: Total likes on TikTok posts featuring the song in 2024.
  - `2024_tik_tok_views`: Total views on TikTok posts featuring the song in 2024.
  - `bpm`: Beats per minute, a measure of song tempo.
  - `key`: Key of the song.
  - `mode`: Mode of the song (major or minor).
  - `danceability`: Percentage indicating how suitable the song is for dancing.
  - `valence`: Positivity of the song's musical content.
  - `energy`: Perceived energy level of the song.
  - `acousticness`: Amount of acoustic sound in the song.
  - `liveness`: Presence of live performance elements.
  - `speechiness`: Amount of spoken words in the song.


The merged 2023/2024 longer dataset (spotify_longer_df) includes the following variables:

- `track`: Name of the song.
- `artist`: Name of the artist(s) of the song.
- `released_year`: Year the song was released.
- `released_month`: Month the song was released.
- `released_day`: Day of the month the song was released.
- `year`: year (2023 or 2024) corresponding to the listed `metric` variable
- `metric`: spotify_streams, spotify_playlist_count, apple_music_playlist_count, or deezer_playlist_count
- `value`: numeric value of the `metric` variable listed
- `bpm`: Beats per minute, a measure of song tempo.
- `key`: Key of the song.
- `mode`: Mode of the song (major or minor).
- `danceability`: Percentage indicating how suitable the song is for dancing.
- `valence`: Positivity of the song's musical content.
- `energy`: Perceived energy level of the song.
- `acousticness`: Amount of acoustic sound in the song.
- `liveness`: Presence of live performance elements.
- `speechiness`: Amount of spoken words in the song.

### Summary Data

*Top 10 Songs by Spotify Streams*
```{r}
top10streams_df <- spotify_df |> 
  mutate(total_streams = `2023_spotify_streams` + `2024_spotify_streams`) |> 
  distinct(track, artist, .keep_all = TRUE) |> 
  arrange(desc(total_streams)) |> 
  slice_head(n = 10)

table_top10streams <- top10streams_df |> 
  select(track, artist, total_streams) |> 
  gt() |> 
  tab_header(
    title = "Top 10 Streamed Songs on Spotify (2023 & 2024)"
  ) |> 
  fmt_number(
    columns = total_streams,
    decimals = 0
  ) |> 
  cols_label(
    track = "Song Title",
    artist = "Artist",
    total_streams = "Streams"
  )

table_top10streams
```
It is important to note that this table shows the top 10 streamed songs on Spotify across 2023 and 2024, but only includes songs that were present in both datasets. Therefore, songs released in 2024 are not included in this summary.

*Top 10 Songs by Presence on Playlists*
```{r}
playlists_df <- spotify_df |> 
  mutate(total_playlists = 
           `2023_spotify_playlist_count` + `2024_spotify_playlist_count` +
           `2023_apple_music_playlist_count` + `2024_apple_music_playlist_count` +
           `2023_deezer_playlist_count` + `2024_deezer_playlist_count`) |> 
  distinct(track, artist, .keep_all = TRUE) |> 
  arrange(desc(total_playlists)) |> 
  slice_head(n = 10)

table_top10playlist <- playlists_df |> 
  select(track, artist, total_playlists) |> 
  gt() |> 
  tab_header(
    title = "Top 10 Songs by Total Playlist Count (2023 & 2024)"
  ) |> 
  fmt_number(
    columns = total_playlists,
    decimals = 0
  ) |> 
  cols_label(
    track = "Song Title",
    artist = "Artist",
    total_playlists = "Total Playlists"
  )

table_top10playlist
```
The total number of playlists that each song is present on includes counts from Spotify, Apple Music, and Deezer. The same limitations apply as discussed in the previous table.

*Top 10 Songs by TikTok Engagement*
```{r}
top10tiktok <- spotify_df |> 
  mutate(tiktok_engagement = 
           `2024_tik_tok_posts` + `2024_tik_tok_likes` + `2024_tik_tok_views`) |> 
  distinct(track, artist, .keep_all = TRUE) |> 
  arrange(desc(tiktok_engagement)) |> 
  slice_head(n = 10)

table_top10tiktok <- top10tiktok |> 
  select(track, artist, tiktok_engagement) |> 
  gt() |> 
  tab_header(
    title = "Top 10 Songs by TikTok Engagement (2024)"
  ) |> 
  fmt_number(
    columns = tiktok_engagement,
    decimals = 0
  ) |> 
  cols_label(
    track = "Song Title",
    artist = "Artist",
    tiktok_engagement = "TikTok Engagement"
  )

table_top10tiktok
```
TikTok engagement was summarized by adding total views, likes, and posts. Data was collected from 2024, but the table only shows songs released before 2024.

*Number of Top Songs By Release Year*
```{r}
songs_by_year <- spotify_df |> 
  count(released_year, name = "song_count") |> 
  arrange(desc(released_year))

table_songs_by_year <- songs_by_year |> 
  gt() |> 
  tab_header(
    title = "Count of Songs by Release Year"
  ) |> 
  fmt_number(
    columns = song_count,
    decimals = 0
  ) |> 
  cols_label(
    released_year = "Release Year",
    song_count = "Number of Songs"
  )

table_songs_by_year
```

An alternative way of displaying this can be via histogram: 

```{r histogram of song release by year}
ggplot(songs_by_year, aes(x = released_year, y = song_count)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  labs(
    title = "Number of Songs by Release Year",
    x = "Release Year",
    y = "Number of Songs"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for better readability
  )
```
This table shows that all top songs consistent across our 2023 and 2023 datasets were released in the year 2000 or later. 2021-2023 show the highest number of top songs, the greatest being in 2022. This may indicate that most songs reach their max popularity when they have had a few years to gain attention and recurring listeners.

### Visualizations and Analysis

*Distribution of BPM by Song Count*
```{r, warning=FALSE}
bpm_histogram <- ggplot(spotify_df, aes(x = bpm)) +
  geom_histogram(stat = "count", binwidth = 5, aes(fill = ..count..), color = "black", alpha = 0.7) +
  scale_fill_viridis_c(option = "plasma", end = 0.9) +
  scale_y_continuous(labels = scales::comma) + 
  labs(
    title = "Histogram of BPM by Song Count",
    subtitle = "Distribution of BPM values across songs in the dataset",
    x = "Beats Per Minute (BPM)",
    y = "Song Count"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

interactive_bpm_histogram <- ggplotly(bpm_histogram)

interactive_bpm_histogram
```

*Distribution of Key by Song Count*
```{r}
key_count_df <- spotify_df |> 
  filter(!is.na(key)) |>
  count(key)

key_histogram <- ggplot(key_count_df, aes(x = reorder(key, n), y = n, fill = n)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.7) +
  scale_fill_viridis_c(option = "plasma", end = 0.9) + 
  labs(
    title = "Histogram of Song Count by Key",
    subtitle = "Distribution of songs across different musical keys",
    x = "Key",
    y = "Song Count"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1))

interactive_key_histogram <- ggplotly(key_histogram)

interactive_key_histogram
```

*Distribution of Mode by Song Count*
```{r}
mode_count_df <- spotify_df |> 
  filter(!is.na(mode)) |>
  count(mode) |> 
  arrange(desc(n))

mode_histogram <- ggplot(mode_count_df, aes(x = reorder(mode, n), y = n, fill = n)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.7) +
  scale_fill_viridis_c(option = "plasma", end = 0.9) +
  labs(
    title = "Histogram of Song Count by Mode",
    subtitle = "Distribution of songs across different musical modes",
    x = "Mode",
    y = "Song Count"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

interactive_mode_histogram <- ggplotly(mode_histogram)

interactive_mode_histogram
```

*Association Between Danceability and TikTok Engagement*
```{r danceability and tiktok enagagement}
dance_df=
  spotify_df|>
  filter(!is.na(danceability))|>
  filter(released_year == "2023")|>
  mutate(tiktok_engagement = 
           `2024_tik_tok_posts` + `2024_tik_tok_likes` + `2024_tik_tok_views`) |> 
  distinct(track, artist, .keep_all = TRUE) |> 
  arrange(desc(tiktok_engagement)) |> 
  group_by(danceability) |> 
  summarise(song_count = n(), tiktok_engagement = mean(tiktok_engagement, na.rm = TRUE))

dance_barplot =
  ggplot(dance_df, aes(x = danceability, y = tiktok_engagement, fill = song_count)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.8) +
  scale_fill_gradient(low = "lightblue", high = "blue") + 
  labs(
    title = "TikTok Engagement by Danceability",
    subtitle = "Bars filled based on song count",
    x = "Danceability",
    y = "TikTok Engagement",
    fill = "Song Count"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) 

interactive_dance_barplot = ggplotly(dance_barplot)

interactive_dance_barplot
```

This plot demonstrates the association between danceability and average TikTok engagement, a variable which is aggregated by posts made with the song, likes and views. The highest engagement on TikTok is among from 4 songs with a danceability score of 71 and the distribution is a weak bimodal one.

*Energy and TikTok Engagement*
```{r energy and tiktok engagement}
energy_df = 
  spotify_df |>
  filter(!is.na(energy)) |>
  filter(released_year == "2023") |>
  mutate(tiktok_engagement = 
           `2024_tik_tok_posts` + `2024_tik_tok_likes` + `2024_tik_tok_views`) |>
  distinct(track, artist, .keep_all = TRUE) |>
  arrange(desc(tiktok_engagement))

energy_scatterplot =
  ggplot(energy_df, aes(x = energy, y = tiktok_engagement)) +
  geom_point(alpha = 0.7, color = "blue") +
  labs(
    title = "TikTok Engagement by Energy, 2023",
    subtitle = "Each point represents a single song",
    x = "Energy",
    y = "TikTok Engagement"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

interactive_energy_scatterplot = ggplotly(energy_scatterplot)

interactive_energy_scatterplot

```

This plot demonstrates the association between energy and average TikTok engagement, a variable which is aggregated by posts made with the song, likes and views. The highest engagement on TikTok is among from 1 song with a energy score of 68.


*Liveness and TikTok Engagement*
```{r liveness and tiktok engagement}
liveness_df = 
  spotify_df |>
  filter(!is.na(liveness)) |>
  filter(released_year == "2023") |>
  mutate(tiktok_engagement = 
           `2024_tik_tok_posts` + `2024_tik_tok_likes` + `2024_tik_tok_views`) |>
  distinct(track, artist, .keep_all = TRUE) |>
  arrange(desc(tiktok_engagement))

# Scatter plot
liveness_scatterplot =
  ggplot(liveness_df, aes(x = liveness, y = tiktok_engagement)) +
  geom_point(alpha = 0.7, color = "blue") +
  labs(
    title = "TikTok Engagement by Liveness",
    subtitle = "Each point represents a single song",
    x = "Liveness",
    y = "TikTok Engagement"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

# Convert to interactive plot
interactive_liveness_scatterplot = ggplotly(liveness_scatterplot)

interactive_liveness_scatterplot
```

This plot demonstrates the association between liveness and average TikTok engagement, a variable which is aggregated by posts made with the song, likes and views. The highest engagement on TikTok is among from 1 song with a liveness score of 3.


*Regression Model of Danceability, Energy, Liveness, Valence, and Acousticness on TikTok Engagement*

```{r regression on tik tok engagement and qualitative features}
spotify_df=
  spotify_df|>
  mutate(tiktok_engagement = 
           `2024_tik_tok_posts` + `2024_tik_tok_likes` + `2024_tik_tok_views`) 

linear_model =
  lm(tiktok_engagement ~ danceability + energy + liveness, data = spotify_df)

tidy_model =
  broom::tidy(linear_model)|> 
  mutate(
    estimate = round(estimate, 3), 
    std.error = round(std.error, 3),
    statistic = round(statistic, 2),
    p.value = ifelse(p.value < 0.001, "<0.001", round(p.value, 3)))

model_table =
  tidy_model |> 
  kable(
    col.names = c("Predictor", "Estimate", "Standard Error", "t-Statistic", "P-Value"),
    caption = "Linear Regression: Qualitative Song Features on TikTok Engagement",
    digits = 3,
    format = "html"
  ) |> 
  kable_styling(
    full_width = FALSE,
    position = "center",
    font_size = 14
  ) 

model_table
```

Based on the linear regression assessing the relationship between TikTok engagement and qualitative features of the song such as danceability, energy, and liveness, none of the variables have a statistically significant association with  TikTok engagement. However, one interesting insight is that danceability has a positive association with TikTok engagement although it is not statistically significant. This regression was driven by the hypothesis that livelier and energetic music will have more TikTok engagement considering the popularity of TikTok dance challenges. Although the findings were not statically significant, they were interesting. 

