---
title: "2023"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

```{r include = FALSE}
library(tidyverse)
library(plotly)
```

# Exploratory Data Analysis
### Discussion of Raw Data

The spotify-2023.csv dataset contains information about the most streamed songs on Spotify in 2023. Each observation includes basic information such as the track name, artist name, release date, and number of streams, as well as additional characteristics like bpm, key, mode, and other audio features. Popularity is measured by presence in Spotify playlists and charts, and across playlists and charts of other streaming platforms including Apple, Deezer, and Shazam.

```{r echo = TRUE, results = 'hide'}
spotify2023_df = 
  read_csv("data/spotify-2023.csv") |> 
  janitor::clean_names() |> 
  rename_with(~ gsub("_percent", "", .), contains("_percent")) |> 
  rename(artist = artist_s_name)

view(spotify2023_df)
```

(Include issues that come up with data here).

The dataset includes the following variables:
- `track_name`: Name of the song.
- `artist`: Name of the artist(s).
- `artist_count`: Number of artists contributing to the song.
- `released_year`: Year the song was released.
- `released_month`: Month the song was released.
- `released_day`: Day of the month the song was released.
- `in_spotify_playlists`: Number of Spotify playlists the song is included in.
- `in_spotify_charts`: Presence and rank of the song on Spotify charts.
- `streams`: Total number of streams on Spotify.
- `in_apple_playlists`: Number of Apple playlists the song is included in.
- `in_apple_charts`: Presence and rank of the song on Apple charts.
- `in_deezer_playlists`: Number of Deezer playlists the song is included in.
- `in_deezer_charts`: Presence and rank of the song on Deezer charts.
- `in_shazam_charts`: Presence and rank of the song on Shazam charts.
- `bpm`: Beats per minute, a measure of song tempo.
- `key`: Key of the song.
- `mode`: Mode of the song (major or minor).
- `danceability`: Percentage indicating how suitable the song is for dancing.
- `valence`: Positivity of the song's musical content.
- `energy`: Perceived energy level of the song.
- `acousticness`: Amount of acoustic sound in the song.
- `instrumentalness`: Amount of instrumental content in the song.
- `liveness`: Presence of live performance elements.
- `speechiness`: Amount of spoken words in the song.

### Summary Data

*Summary Statistics*

```{r}
summary_table = 
  summarize(
    spotify_streams_mean = mean(spotify_streams, na.rm = TRUE),
    spotify_streams_median = median(spotify_streams, na.rm = TRUE),
    spotify_streams_iqr = IQR(spotify_streams, na.rm = TRUE),
    
    youtube_views_mean = mean(you_tube_views, na.rm = TRUE),
    youtube_views_median = median(you_tube_views, na.rm = TRUE),
    youtube_views_iqr = IQR(you_tube_views, na.rm = TRUE),
    
    soundcloud_streams_mean = mean(soundcloud_streams, na.rm = TRUE),
    soundcloud_streams_median = median(soundcloud_streams, na.rm = TRUE),
    soundcloud_streams_iqr = IQR(soundcloud_streams, na.rm = TRUE),
    
    tiktok_views_mean = mean(tik_tok_views, na.rm = TRUE),
    tiktok_views_median = median(tik_tok_views, na.rm = TRUE),
    tiktok_views_iqr = IQR(tik_tok_views, na.rm = TRUE)
  )
```


=======
This will be the rmd for 2023 analysis...
>>>>>>> a7a6440417c688ea335b86a025c3ebb9866b8ef9
