---
title: "2023 and 2024"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide 
editor_options: 
  chunk_output_type: inline
---
```{r include=FALSE}
library(tidyverse)
library(plotly)
library(ggplot2)
library(gt)
```
# Exploratory Data Analysis
### Discussion of Raw Data

INSERT INFO ABOUT 2023 and 2024 data

```{r echo=TRUE, include=FALSE}
spotify2024_df= read_csv("data/2024_data.csv") |>
                janitor::clean_names() |>
                select(-tidal_popularity)|>
  mutate(
    track = iconv(track, from = "UTF-8", to = "ASCII//TRANSLIT"),
    artist = iconv(artist, from = "UTF-8", to = "ASCII//TRANSLIT")
  ) |> 
  filter(artist!= "xSyborg")

spotify2023_df = 
  read_csv("data/spotify-2023.csv") |> 
  janitor::clean_names() |> 
  rename_with(~ gsub("_percent", "", .), contains("_percent")) |> 
  rename(artist = artist_s_name) |> 
  mutate(streams = ifelse(track_name == "Love Grows (Where My Rosemary Goes)", NA, streams),
         streams = as.numeric(streams)) |> 
  rename(track=track_name)


```

TALK ABOUT CHANGES MADE TO DATASETS PRIOR TO MERGING

The merged 2023/2024 dataset includes the following variables (CHANGE LIST):

- `track_name`: Name of the song.
- `album_name`: Name of the album the song belongs to.
- `artist`: Name of the artist(s) of the song.
- `release_date`: Date when the song was released.
- `isrc`: International Standard Recording Code for the song.
- `all_time_rank`: Ranking of the song based on its all-time popularity.
- `track_score`: Score assigned to the track based on various factors.
- `spotify_streams`: Total number of streams on Spotify.
- `spotify_playlist_count`: Number of Spotify playlists the song is included in.
- `spotify_playlist_reach`: Reach of the song across Spotify playlists.
- `spotify_popularity`: Popularity score of the song on Spotify.
- `you_tube_views`: Total views of the song's official video on YouTube.
- `you_tube_likes`: Total likes on the song's official video on YouTube.
- `tik_tok_posts`: Number of TikTok posts featuring the song.
- `tik_tok_likes`: Total likes on TikTok posts featuring the song.
- `tik_tok_views`: Total views on TikTok posts featuring the song.
- `you_tube_playlist_reach`: Reach of the song across YouTube playlists.
- `apple_music_playlist_count`: Number of Apple Music playlists the song is included in.
- `airplay_spins`: Number of times the song has been played on radio stations.
- `siriusxm_spins`: Number of times the song has been played on SiriusXM.
- `deezer_playlist_count`: Number of Deezer playlists the song is included in.
- `deezer_playlist_reach`: Reach of the song across Deezer playlists.
- `amazon_playlist_count`: Number of Amazon Music playlists the song is included in.
- `pandora_streams`: Total number of streams on Pandora.
- `pandora_track_stations`: Number of Pandora stations featuring the song.
- `soundcloud_streams`: Total number of streams on Soundcloud.
- `shazam_counts`: Total number of times the song has been Shazamed.
- `explicit_track`: Indicates whether the song contains explicit content.

### Summary Data

### Visualizations and Analysis

