---
title: "2024"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide 
editor_options: 
  chunk_output_type: console
---
```{r include=FALSE}
library(readr)
library(janitor)
library(dplyr)
library(ggplot2)
library(gt)
library(tidyr)
```
### Exploratory Data Analysis

### Discussion of Raw Data

The 2024_data.csv dataset contains information about the most streamed songs on Spotify in 2024. Each track includes general basic information like the track name, artist name, and album name as well as extensive information on its reach across several streaming platforms (including Spotify, SoundCloud, Youtube, TikTok, Apple Music, etc. ). Popularity is measured across these different platforms through variables such as all time rank, number of streams, playlist reach, and playlist count.

```{r echo=TRUE, include=FALSE}
dataset_2024 <- read_csv("data/2024_data.csv") |>
                clean_names() |>
                select(-tidal_popularity)|>
  mutate(
    track = iconv(track, from = "UTF-8", to = "ASCII//TRANSLIT"),
    artist = iconv(artist, from = "UTF-8", to = "ASCII//TRANSLIT")
  ) |> 
  filter(artist!= "xSyborg")
```

Some issues that came up was with values having unsupported characters (e.g., in the album_name column) and that did not allow for analysis for the all time rank analysis. Another issue that we filtered out was that two tracks had a duplicate observation, but the duplicate observations did not have any streaming information and both had "xSyborg" listed as the artist name.

The dataset includes the following variables:

- `track_name`: Name of the song.
- `album_name`: Name of the album the song belongs to.
- `artist`: Name of the artist(s) of the song.
- `release_date`: Date when the song was released.
- `isrc`: International Standard Recording Code for the song.
- `all_time_rank`: Ranking of the song based on its all-time popularity.
- `track_score`: Score assigned to the track based on various factors.
- `spotify_streams`: Total number of streams on Spotify.
- `spotify_playlist_count`: Number of Spotify playlists the song is included in.
- `spotify_playlist_reach`: Reach of the song across Spotify playlists.
- `spotify_popularity`: Popularity score of the song on Spotify.
- `you_tube_views`: Total views of the song's official video on YouTube.
- `you_tube_likes`: Total likes on the song's official video on YouTube.
- `tik_tok_posts`: Number of TikTok posts featuring the song.
- `tik_tok_likes`: Total likes on TikTok posts featuring the song.
- `tik_tok_views`: Total views on TikTok posts featuring the song.
- `you_tube_playlist_reach`: Reach of the song across YouTube playlists.
- `apple_music_playlist_count`: Number of Apple Music playlists the song is included in.
- `airplay_spins`: Number of times the song has been played on radio stations.
- `siriusxm_spins`: Number of times the song has been played on SiriusXM.
- `deezer_playlist_count`: Number of Deezer playlists the song is included in.
- `deezer_playlist_reach`: Reach of the song across Deezer playlists.
- `amazon_playlist_count`: Number of Amazon Music playlists the song is included in.
- `pandora_streams`: Total number of streams on Pandora.
- `pandora_track_stations`: Number of Pandora stations featuring the song.
- `soundcloud_streams`: Total number of streams on Soundcloud.
- `shazam_counts`: Total number of times the song has been Shazamed.
- `explicit_track`: Indicates whether the song contains explicit content.

# 2024 Most Streamed Songs on Spotify 

### Summary Statistics
```{r}
summary_table=
dataset_2024 |> 
  summarize(
    spotify_streams_mean = mean(spotify_streams, na.rm = TRUE),
    spotify_streams_median = median(spotify_streams, na.rm = TRUE),
    spotify_streams_iqr = IQR(spotify_streams, na.rm = TRUE),
    
    youtube_views_mean = mean(you_tube_views, na.rm = TRUE),
    youtube_views_median = median(you_tube_views, na.rm = TRUE),
    youtube_views_iqr = IQR(you_tube_views, na.rm = TRUE),
    
    soundcloud_streams_mean = mean(soundcloud_streams, na.rm = TRUE),
    soundcloud_streams_median = median(soundcloud_streams, na.rm = TRUE),
    soundcloud_streams_iqr = IQR(soundcloud_streams, na.rm = TRUE),
    
    tiktok_views_mean = mean(tik_tok_views, na.rm = TRUE),
    tiktok_views_median = median(tik_tok_views, na.rm = TRUE),
    tiktok_views_iqr = IQR(tik_tok_views, na.rm = TRUE)
  ) |> 
  pivot_longer(
    cols = everything(),
    names_to = c("Platform", "Statistic"),  
    names_pattern = "(.*)_(.*)"             
  ) |> 
  mutate(
    Platform = case_when(
      Platform == "spotify_streams" ~ "Spotify Streams",
      Platform == "youtube_views" ~ "YouTube Views",
      Platform == "soundcloud_streams" ~ "SoundCloud Streams",
      Platform == "tiktok_views" ~ "TikTok Views",
      TRUE ~ Platform  
    )
  ) |> 
  pivot_wider(
    names_from = Statistic,               
    values_from = value
  )

summary_table |> 
  gt() |> 
  tab_header(
    title = "Summary Statistics Across Platforms",
    subtitle = "Mean, Median, and Interquartile Range"
  ) |> 
  cols_label(
    Platform = "Platform",
    mean = "Mean",
    median = "Median",
    iqr = "Interquartile Range"
  ) |> 
  fmt_number(
    columns = c(mean, median, iqr),
    decimals = 0
  ) |> 
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels()
  )

```
*Interpretation*
The mean versus median streams or views across all the platforms are very different from each other, indicating that the distribution of data is skewed for each platform. It is also interesting to note TikTok's popularity, where the platform has both the most mean and median views compared to the other platforms. Spotify has the second greatest amount of mean and median streams for 2024.

### Top 10 Songs All-Time Rank

```{r all time rank}
top10rank_songs=
  dataset_2024 |>
  slice_head(n = 10)

table_top10rank =
  top10rank_songs |>
  select(track, artist, all_time_rank) |>
         gt() |>
         tab_header(
           title = "Top 10 All Time Rank in 2024") |>
         fmt_number(
           columns = all_time_rank,
           decimals = 0
         ) |>
         cols_label(
           track = "Song Title",
           artist = "Artist",
           all_time_rank = "All Time Rank"
         )

table_top10rank
```

Interpretation: This table displays the songs and artists for 2024 based on the all time rank, a variable in the data set that captures all time popularity. There are no repeated artists in this list meaning no single artist had more than one song in the top 10 all time rank.  

### Percentage of Explicit Tracks

```{r percent explicit, warning=FALSE}
explicit_song =
  dataset_2024 |>
  mutate(explicit_track = ifelse(explicit_track == 1, "Explicit", "Non-Explicit"))|>
  count(explicit_track) |>
  mutate(percentage = n / sum(n) * 100)
  
  explicit_pie_chart =
  ggplot(explicit_song, aes(x = "", y = percentage, fill = explicit_track)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") + 
  geom_text(
    aes(label = paste0(round(percentage, 1), "%")), 
    position = position_stack(vjust = 0.5),
    color = "white", 
    size = 5 ) +
  labs(
    title = "Percentage of Explicit vs Non-Explicit Songs",
    fill = "Song Type") +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")) +
  scale_fill_manual(values = c("Explicit" = "#E63946", "Non-Explicit" = "#457B9D"))
  
  explicit_pie_chart
```

Interpretation: This visualization is representing the percentage tracks categorized as explicit and not explicit in the entire data set. Evidently more than half of the songs are non-ecplicit (63.7%). 

## Top 10 Artists by Different Media Platforms

### Top 10 Artists by Number of Most Streamed Songs

```{r top artists}
data_set_artist =
 dataset_2024 |>
    drop_na(artist)|>
  group_by(artist) |>
  summarise(
    song_count = n(),  
    .groups = "drop")

top10artist_songs=
  data_set_artist |>
  arrange(desc(song_count)) |>
  slice_head(n = 10)

table_top10artist_songs =
  top10artist_songs |>
  select(artist, song_count) |>
         gt() |>
         tab_header(
           title = "Top 10 Artists in 2024") |>
         fmt_number(
           columns = song_count,
           decimals = 0
         ) |>
         cols_label(
           artist = "Artist",
           song_count = "Song Count"
         )

table_top10artist_songs
```

Interpretation: This table represent the top 10 artists in the data set based on the amount of songs included in the data set which captures the most streamed songs. The mix of artists is diverse including Latin music, reggeton, pop, and rap artists. 

### Artists with the Top 10 Spotify Streams

```{r}
 dataset_2024 |>
    drop_na(artist)|>
  group_by(artist) |>
  summarize(total_spotify_streams = sum(spotify_streams, na.rm = TRUE)) |> 
  arrange(desc(total_spotify_streams)) |> 
  top_n(10, total_spotify_streams) |> 
  ggplot(aes(x=total_spotify_streams, y=reorder(artist, total_spotify_streams))) +
  geom_bar(stat = "identity", fill = "skyblue") + 
  labs(
    title = "Top 10 Artists Spotify Streams",
    x = "Total Spotify Streams",
    y = "Artist"
  ) +
  theme_minimal()
  
```

### Artists with the Top 10 Soundcloud Streams
```{r}
 dataset_2024 |>
    drop_na(artist)|>
  group_by(artist) |>
  summarize(total_sc_streams = sum(soundcloud_streams, na.rm = TRUE)) |> 
  arrange(desc(total_sc_streams)) |> 
  top_n(10, total_sc_streams) |> 
  ggplot(aes(x=total_sc_streams, y=reorder(artist, total_sc_streams))) +
  geom_bar(stat = "identity", fill = "skyblue") + 
  labs(
    title = "Top 10 Artists Soundcloud Streams",
    x = "Total Soundcloud Streams",
    y = "Artist"
  ) +
  theme_minimal()
```

### Artists with the Top 10 YouTube Views
```{r}
 dataset_2024 |>
    drop_na(artist)|>
  group_by(artist) |>
  summarize(total_yt = sum(you_tube_views, na.rm = TRUE)) |> 
  arrange(desc(total_yt)) |> 
  top_n(10, total_yt) |> 
  ggplot(aes(x=total_yt, y=reorder(artist, total_yt))) +
  geom_bar(stat = "identity", fill = "skyblue") + 
  labs(
    title = "Top 10 Artists YouTube Views",
    x = "Total YouTube Views",
    y = "Artist"
  ) +
  theme_minimal()
```

### Artists with the Top 10 TikTok Views
```{r}
 dataset_2024 |>
    drop_na(artist)|>
  group_by(artist) |>
  summarize(total_tt = sum(tik_tok_views, na.rm = TRUE)) |> 
  arrange(desc(total_tt)) |> 
  top_n(10, total_tt) |> 
  ggplot(aes(x=total_tt, y=reorder(artist, total_tt))) +
  geom_bar(stat = "identity", fill = "skyblue") + 
  labs(
    title = "Top 10 Artists TikTok Views",
    x = "Total TikTok Views",
    y = "Artist"
  ) +
  theme_minimal()
```

*Interpretation*
The top 10 artists with the most streams or views across Spotify, Soundcloud, YouTube, and TikTok vary widely. Streaming platforms like Spotify and TikTok (which have the highest number of user streams) have top artists that more fall under the pop genre, while Soundcloud has almost exclusively rap artists (with the exception of Major Lazer). 

## Top Songs on Streaming Apps

### Top 10 Streamed Songs
```{r top 10 streams}
dataset_2024 =
  dataset_2024 |>
  mutate(
    cumulative_streams = rowSums(across(c(pandora_streams, spotify_streams, soundcloud_streams, you_tube_views)), na.rm = TRUE)
  )

top10stream_songs=
  dataset_2024 |>
  arrange(desc(cumulative_streams)) |>
  slice_head(n = 10)

table_top10streams =
  top10stream_songs |>
  select(track, artist, cumulative_streams) |>
         gt() |>
         tab_header(
           title = "Top 10 Streamed Songs in 2024",
           subtitle = "Aggregated across Spotify, Pandora, YouTube, TikTok, and Soundcloud"
         ) |>
         fmt_number(
           columns = cumulative_streams,
           decimals = 0
         ) |>
         cols_label(
           track = "Song Title",
           artist = "Artist",
           cumulative_streams = "Cumulative Streams"
         )

table_top10streams
```

*Interpretation*
This table aggregates the views and streams from Spotify, Pandora, YouTube and SoundCloud to show which song had the most streams. Streaming is defined loosely to include views as streams, both of which are arguably passive consumption compared to a like. The highest streamed song has over 200 billion streams. There is no overlap between the top 10 songs by all time rank and the list of most streamed songs indicating some dissonance between the the popularity of the song as defined by the all_time_rank variable and the streams.  


### Top 10 Streamed Songs on Spotify and Soundcloud

```{r}
get_top_songs <- function(data, metric, top_n = 10) {
  data |> 
    select(track, artist, !!sym(metric)) |> 
    drop_na(track) |> 
    arrange(desc(!!sym(metric))) |> 
    head(top_n) |> 
    gt() |> 
    tab_header(
      title = paste("Top Songs by", gsub("_", " ", metric)),
      subtitle = paste("Based on", metric, "metrics")
    ) |> 
    fmt_number(
      columns = !!sym(metric),
      decimals = 0
    ) |> 
    cols_label(
      track = "Track Name",
      artist = "Artist",
      !!sym(metric) := gsub("_", " ", metric)
    )
}
top_spotify <- get_top_songs(dataset_2024, "spotify_streams")
top_soundcloud <- get_top_songs(dataset_2024, "soundcloud_streams")
top_youtube <- get_top_songs(dataset_2024, "you_tube_views")
top_tiktok <- get_top_songs(dataset_2024, "tik_tok_views")

top_spotify
top_soundcloud
```
*Interpretation* 
Interestingly, the top 10 spotify streamed songs and the top 10 soundcloud streamed songs had no overlap in track or artist. Spotify's top songs are more mainstream and fall under the Pop genre, while SoundCloud's top songs fall under the Hip Hop/Rap genres.

## Top Songs on Socials

###Top 10 Liked Songs (Based on TikTok Likes and TikTok Views)

```{r top 10 tiktok}
top10tiktok_songs=
  dataset_2024 |>
  arrange(desc(tik_tok_likes)) |>
  slice_head(n = 10)

table_top10tiktok =
  top10tiktok_songs |>
  select(track, artist, tik_tok_likes) |>
         gt() |>
         tab_header(
           title = "Top 10 Songs by TikTok Likes"
         ) |>
         fmt_number(
           columns = tik_tok_likes,
           decimals = 0
         ) |>
         cols_label(
           track = "Song Title",
           artist = "Artist",
           tik_tok_likes = "TikTok Likes"
         )

table_top10tiktok
top_tiktok
```

*Interpretation* 
In the table capturing top 10 songs by count of TikTok likes, there is overlap between the most streamed songs specifically "Monkeys Spinning Monkeys", "Love You So", "Oh No", "Funny Song", "Aesthetic", "STAY (with Justin Bieber)", and "Pieces". This makes sense considering TikTok views might in the aggregation of streams skew the data in the previous table. However, it might also point to a greater phenomena recognized on social media as the "TikTok-ification" of music, where popularity and excellence might be defined by streams, likes, and 'going viral' on TikTok. In the table capturing top 10 songs by count of TikTok views, we see that the top three songs are the same as the top four songs for TikTok likes, and "STAY (with Justin Bieber)" and "Pieces" are ranked #10 and #8 respectively for both charts as well. Other songs that are  included in both charts, but are ranked differently, include "Spongebob", "Funny Song", and "She Share Story",  

### Top 10 Songs (Based on YouTube Likes and YouTube Views)

```{r top 10 youtube}
top10yt_songs=
  dataset_2024 |>
  arrange(desc(you_tube_likes)) |>
  slice_head(n = 10)

table_top10yt =
  top10yt_songs |>
  select(track, artist, you_tube_likes) |>
         gt() |>
         tab_header(
           title = "Top 10 Songs by YouTube Likes"
         ) |>
         fmt_number(
           columns = you_tube_likes,
           decimals = 0
         ) |>
         cols_label(
           track = "Song Title",
           artist = "Artist",
           you_tube_likes = "YouTube Likes"
         )

table_top10yt
top_youtube
```

Interpretation: The table capturing the top 10 songs by YouTube likes and offers a different insight on song popularity. YouTube is a platform mostly for music videos so this captures a different measure for song popularity based on the music video popularity. This list does not have any overlap with the songs from the previous tables. One notable song in this list is Baby Shark which is toop 3 with most youtube likes in the list. The table displaying the top 10 songs by YouTube views also includes "Despacito" as its #2 ranked song. Other songs that are in both tables, but are ranked differently, include "Baby Shark", "Shape of You", "See You Again (feat. Charlie Puth)", and "Perfect".

### Streaming Platform Data

```{r}
get_top_5_songs <- function(data, metric, top_n = 5) {
  data |> 
    select(track, artist, !!sym(metric)) |> 
    drop_na(track) |> 
    arrange(desc(!!sym(metric))) |> 
    head(top_n) |> 
    gt() |> 
    tab_header(
      title = paste("Top Songs by", gsub("_", " ", metric)),
      subtitle = paste("Based on", metric, "metrics")
    ) |> 
    fmt_number(
      columns = !!sym(metric),
      decimals = 0
    ) |> 
    cols_label(
      track = "Track Name",
      artist = "Artist",
      !!sym(metric) := gsub("_", " ", metric)
    )
}
top_5_spotify <- get_top_5_songs(dataset_2024, "spotify_streams")
top_5_soundcloud <- get_top_5_songs(dataset_2024, "soundcloud_streams")
top_5_youtube <- get_top_5_songs(dataset_2024, "you_tube_views")
top_5_tiktok <- get_top_5_songs(dataset_2024, "tik_tok_views")

top_5_spotify
top_5_soundcloud
top_5_youtube
top_5_tiktok
```

*Interpretation*
Based on this 2024 dataset, there is almost no overlap among the most popular five songs streamed for Spotify, Soundcloud, YouTube, and TikTok. The only track in common is "Shape of You" by Ed Sheeran, where this is on the Top 5 Songs for YouTube and Spotify. These tables also allow us to compare popularity of streaming platforms, where Tik Tok's most viewed song (Mokeys Spinning Monkeys by Kevin MacLeod) has 233,232,311,463 views, and Soundcloud's most streamed song has 319,835,943 streams. This factor of a hundred difference demonstrates TikTok's popularity compared to other streaming platforms such as SoundCloud. It is also interesting that SoundCloud's Top 5 streamed songs are all in the rap genre, where Spotify's Top 5 streamed songs generally fall into today's hits/pop. 


## Streams, Views, and Intrigue by TikTok Popularity

### Comparing Tik Tok Likes to Streams Across Spotify, Soundcloud, and Youtube

```{r, warning=FALSE}
dataset_2024 |> 
  drop_na(tik_tok_likes) |> 
  mutate(tik_tok_likes_log=log10(tik_tok_likes)) |> 
mutate(
    tiktok_category = case_when(
      tik_tok_likes_log <= log10(1e6) ~ "Low",       
      tik_tok_likes_log <= log10(1e8) ~ "Medium",    
      tik_tok_likes_log > log10(1e8) ~ "High"        
    ),
    tiktok_category = factor(tiktok_category, levels = c("Low", "Medium", "High")) 
  ) |> 
  ggplot(aes(x = tiktok_category, y = spotify_streams, fill = tiktok_category)) +
  geom_boxplot() +
  labs(
    title = "Comparing Spotify Streams to Amount of TikTok Likes",
    x = "TikTok Likes",
    y = "Spotify Streams",
   fill = "TikTok Likes Category"
  ) +
  theme_minimal()


dataset_2024 |> 
  drop_na(tik_tok_likes) |> 
  mutate(tik_tok_likes_log=log10(tik_tok_likes)) |> 
mutate(
    tiktok_category = case_when(
      tik_tok_likes_log <= log10(1e6) ~ "Low",       
      tik_tok_likes_log <= log10(1e8) ~ "Medium",    
      tik_tok_likes_log > log10(1e8) ~ "High"        
    ),
    tiktok_category = factor(tiktok_category, levels = c("Low", "Medium", "High")) 
  ) |> 
  ggplot(aes(x = tiktok_category, y = you_tube_views, fill = tiktok_category)) +
  geom_boxplot() +
  labs(
    title = "Comparing YouTube Views to Amount of TikTok Likes",
    x = "TikTok Likes",
    y = "Log Base 10 of YouTube Views",
   fill = "TikTok Likes Category"
  ) +
  scale_y_log10() +
  theme_minimal()

dataset_2024 |> 
  drop_na(tik_tok_likes) |> 
  mutate(tik_tok_likes_log=log10(tik_tok_likes)) |> 
mutate(
    tiktok_category = case_when(
      tik_tok_likes_log <= log10(1e6) ~ "Low",       
      tik_tok_likes_log <= log10(1e8) ~ "Medium",    
      tik_tok_likes_log > log10(1e8) ~ "High"        
    ),
    tiktok_category = factor(tiktok_category, levels = c("Low", "Medium", "High")) 
  ) |> 
  ggplot(aes(x = tiktok_category, y = soundcloud_streams, fill = tiktok_category)) +
  geom_boxplot() +
  labs(
    title = "Comparing Soundcloud Streams to Amount of TikTok Likes",
    x = "TikTok Likes",
    y = "Log Base 10 of Soundcloud Streams",
   fill = "TikTok Likes Category"
  ) +
  scale_y_log10() +
  theme_minimal()
```

### Comparing Tik Tok Likes to Shazam Counts

```{r}
dataset_2024 |> 
  drop_na(tik_tok_likes) |> 
  mutate(tik_tok_likes_log=log10(tik_tok_likes)) |> 
mutate(
    tiktok_category = case_when(
      tik_tok_likes_log <= log10(1e6) ~ "Low",       
      tik_tok_likes_log <= log10(1e8) ~ "Medium",    
      tik_tok_likes_log > log10(1e8) ~ "High"        
    ),
    tiktok_category = factor(tiktok_category, levels = c("Low", "Medium", "High")) 
  ) |> 
  ggplot(aes(x = tiktok_category, y = shazam_counts, fill = tiktok_category)) +
  geom_boxplot() +
  labs(
    title = "Comparing Shazam Counts to Amount of TikTok Likes",
    x = "TikTok Likes",
    y = "Log Base 10 of Shazam Counts",
   fill = "TikTok Likes Category"
  ) +
  scale_y_log10() +
  theme_minimal()
```

*Interpretation*
When looking at the graphs that compare Spotify streams,  YouTube views, Soundcloud streams, and Shazam counts, one can see that increased Tik Tok likes (going from the Low to High category) is associated with increased track streams or views. This makes sense that the more popular songs on streaming platforms or songs that people want to know the name of (through using Shazam) would also be popular on TikTok. 

Note: Due to the very large range of TikTok likes, a log base 10 was taken when creating a categorical variable (`tiktok_category`) for TikTok likes. Log base 10 was also taken of YouTube Views, Soundcloud streams, and Shazam counts to make data easier to digest and interpret when comparing to TikTok Likes categories.







