---
title: "Decoding Spotify Music Trends"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    code_folding: hide
---

```{r include=FALSE}
library(tidyverse)
library(plotly)
library(ggplot2)
library(gt)
library(readr)
library(janitor)
library(dplyr)
library(tidyr)
```

## Motivation

The initial motivation for the project was to expand the understanding of why songs and artists trend on music streaming platforms, such as Spotify. Quality of music is certainly subjective, however it is interesting to consider whether measurable qualities such as danceability, valence, or beats per minute are associated with popularity. As a step further, we were motivated to examine whether these relationships between measurable quantities and popularities remained consistent throughout the calendar year or rather ebbed and flowed as the seasons changed. 

We were also curious about the consistency and interaction between various streaming and social media platforms. With the rapid rise of the social media site tik tok and continued usage of platforms such as Youtube and SoundCloud, we looked to examine whether popularity on one site might correlate to another, or alternatively if the sites appealed to user bases with different music preferences. 

Overall, our project goal was to provide context for why songs and artists become and maintain popularity. We aimed to perform exploratory analysis on our original motivating questions, as well as any secondary questions that arose throughout the project. 

## Related Work

We found a few sources of inspiration for this project, the first being Spotify Wrapped. Each year, Spotify provides users with a summary of their most listened to songs, artists, genres, etc., giving insight into their preferences. While we each have different music preferences, we thought it interesting to consider whether some qualities of music were universally appealing. 

While beginning our research, we found other projects in this space that served as further inspiration, including a capstone project completed by students at Carnegie Mellon University titled “[Competitive Analysis of the Music Streaming Industry](https://amt-lab.org/blog/2019/9/a-competitive-analysis-of-the-music-streaming-industry)”. This project considered trends and the competitive aspects of the music streaming market, and included consideration of various streaming and social media platforms. This project was influential as we decided to include interaction between various streaming platforms as part of our analysis. 

## Initial Questions

In the 2023 dataset, we were interested in exploring song popularity on Spotify, Apple Music, Deezer, and Shazam to determine if top songs were consistent across different music streaming platforms. We also wanted to know how various song characteristics such as BPM, key, mode, danceability, etc. differed across top songs to find out what makes a hit track.

In the 2024 dataset, we were interested in seeing if song popularity differed across various streaming platforms (e.g., Spotify, Soundcloud, YouTube, TikTok, Apple Music, etc.) as well as if different artists were more popular on certain platforms. As we began to analyze and parse through the data, we also were interested in whether a track being explicit or not would affect its popularity or number of streams. We also wanted to specifically see if TikTok engagement (e.g., TikTok likes, views, posts) was related to the number of streams across various platforms.

By combining the 2023 and 2024 datasets, we hoped to assess overlap in top songs and evaluate common characteristics of top songs, where applicable. We also were interested in seeing if specific track characteristics (e.g., danceability) were associated with greater popularity (e.g., TikTok engagement).

## Data

#### 2023 Data
```{r message = FALSE}

spotify2023_df = 
  read_csv("spotify-2023.csv") |> 
  janitor::clean_names() |> 
  rename_with(~ gsub("_percent", "", .), contains("_percent")) |> 
  rename(artist = artist_s_name)
view(spotify2023_df)

spotify2023_df = 
  read_csv("data/spotify-2023.csv") |> 
  janitor::clean_names() |> 
  rename_with(~ gsub("_percent", "", .), contains("_percent")) |> 
  rename(artist = artist_s_name) |> 
  mutate(streams = ifelse(track_name == "Love Grows (Where My Rosemary Goes)", NA, streams),
         streams = as.numeric(streams))

view(spotify2023_df)
```

The spotify-2023.csv dataset contains information about the most streamed songs on Spotify in 2023. Each observation includes basic information such as track name, artist, and release date, as well as streaming data on various platforms including Spotify, Apple Music, Deezer, and Shazam. Popularity is measured through total streams, presence on playlists, and presence on charts. Additionally, BPM, key, mode, and other music characteristics like danceability, energy, and liveness were included for each of the top tracks.

The 2023 dataset was already fairly clean to begin with, but a few changes were made for clarity and consistency. Variable names were simplified and only one stream observation had to be set as “NA” due to the presence of an uninterpretable character variable rather than a numeric value. There were no duplicates in this dataset.

#### 2024 Data
```{r message=FALSE}
dataset_2024 <- read_csv("data/2024_data.csv") |>
                clean_names() |>
                select(-tidal_popularity)|>
  mutate(
    track = iconv(track, from = "UTF-8", to = "ASCII//TRANSLIT"),
    artist = iconv(artist, from = "UTF-8", to = "ASCII//TRANSLIT")
  ) |> 
  filter(artist!= "xSyborg")
```
The 2024_data.csv dataset contains information about the most streamed songs on Spotify in 2024. Each track includes general basic information like the track name, artist name, and album name as well as extensive information on its reach across several streaming platforms (including Spotify, SoundCloud, Youtube, TikTok, Apple Music, etc. ). Popularity is measured across these different platforms through variables such as all time rank, number of streams, playlist reach, and playlist count.

Some issues that came up while cleaning the data was with values having unsupported characters (e.g., in the album_name column) and that did not allow for analysis for the all time rank analysis. Another issue that we filtered out was that two tracks had a duplicate observation, but the duplicate observations did not have any streaming information and both had "xSyborg" listed as the artist name. The `tidal_popularity` variable was removed because the entire column had no observations.

#### 2023-2024 Cross-Year Data
```{r message=FALSE, echo=TRUE}
spotify2024_df= read_csv("data/2024_data.csv") |>
                janitor::clean_names() |>
                select(-tidal_popularity)|>
  mutate(
    track = iconv(track, from = "UTF-8", to = "ASCII//TRANSLIT"),
    artist = iconv(artist, from = "UTF-8", to = "ASCII//TRANSLIT")
  ) |> 
  separate(
    col = release_date,
    into = c("released_month", "released_day", "released_year"),
    sep = "/",
    convert = TRUE  
  ) |> 
    rename(
    `2024_spotify_streams` = spotify_streams,
    `2024_spotify_playlist_count` = spotify_playlist_count,
    `2024_apple_music_playlist_count` = apple_music_playlist_count,
    `2024_deezer_playlist_count` = deezer_playlist_count,
    `2024_tik_tok_posts`= tik_tok_posts,
     `2024_tik_tok_likes`= tik_tok_likes,
     `2024_tik_tok_views`= tik_tok_views,
  )|> 
  filter(artist!= "xSyborg") |> 
  select(track, released_month, released_day, released_year, `2024_spotify_streams`, `2024_spotify_playlist_count`, `2024_apple_music_playlist_count`,
    `2024_deezer_playlist_count`, `2024_tik_tok_posts`, `2024_tik_tok_likes`, `2024_tik_tok_views`)

spotify2023_df = 
  read_csv("data/spotify-2023.csv") |> 
  janitor::clean_names() |> 
  rename_with(~ gsub("_percent", "", .), contains("_percent")) |> 
  rename(artist = artist_s_name) |> 
  mutate(streams = ifelse(track_name == "Love Grows (Where My Rosemary Goes)", NA, streams),
         streams = as.numeric(streams)) |> 
  rename(
    track=track_name,
    '2023_spotify_streams' = streams,
    '2023_spotify_playlist_count' = in_spotify_playlists,
    '2023_apple_music_playlist_count' = in_apple_playlists,
    '2023_deezer_playlist_count' = in_deezer_playlists
    ) |> 
  select(track, artist, released_month, released_day, released_year, '2023_spotify_streams', '2023_spotify_playlist_count', '2023_apple_music_playlist_count', '2023_deezer_playlist_count', bpm, key, mode, danceability, valence, energy, acousticness, liveness, speechiness )

spotify_df=
  inner_join(spotify2023_df, spotify2024_df, by= c("track", "released_month", "released_day", "released_year"), relationship="many-to-many")

```

Prior to merging the aforementioned datasets, in the 2023 dataset, we renamed  `track_name` to `track`, `streams` to `2023_spotify_streams`, `in_spotify_playlists` to `2023_spotify_playlist_count`, `in_apple_playlists` to `2023_apple_music_playlist_count`, and`in_deezer_playlists` to `2023_deezer_playlist_count`. In the 2024 dataset, we separated `release_date` into `released_year`, `released_month`, `released_day`, changed `spotify_streams` to `2024_spotify_streams`, `spotify_playlist_count`  to `2024_spotify_playlist_count`, `apple_music_playlist_count` to `2024_apple_music_playlist_count`, and `deezer_playlist_count` to `2024_deezer_playlist_count`. Variables that were removed from the 2023 dataset prior to merging include the following: `artist_count`, `in_spotify_charts`, `in_apple_charts`, `in_deezer_charts`, `in_shazam_charts`, and `instrumentalness`. Variables that were removed from the 2024 dataset prior to merging include the following: `album_name`, `isrc`, `all_time_rank`, `track_score`, `spotify_playlist_reach`, `spotify_popularity`, `you_tube_views`, `you_tube_likes`, `air_play_spins`, `sirius_xm_spins`, `deezer_playlist_reach`, `amazon_playlist_count`, `pandora_streams`, and `pandora_track_stations`. 

Datasets were merged based on `track`, `released_year`, `released_month`, and `released_day`. The `artist` variable for the 2023 was kept because the 2023 dataset includes the artists that are featured on the track as well. 

When merging the datasets (through inner_join), we found that there were 17 duplicate songs where the duplicate observations contained information for both 2023 and 2024 Spotify streams, Spotify playlist count, and Apple Music playlist count. Some duplicates, however, have varying missing information about 2024 Deezer playlist count and/or TikTok data. Some other notable discrepancies we found among the 17 tracks were the following:
  * “Daylight” and “Numb” are duplicate track names but are from different artists 
  * “My Universe” has limited information in both duplicates
  * “SNAP” has two duplicates (4 total) 
  * “SPIT IN MY FACE” is a duplicate but they are in different keys/have different song details (e.g., danceability, valence, etc.)

## Exploratory Analysis

#### 2023 Exploratory Analysis

*Tuning Into Music: A Journey Through Keys and Modes*

The exploration of musical keys and modes offers valuable insights into the tonal character of songs that resonated the most with listeners in 2023. Through the distribution of Spotify streams and the count of songs in different musical keys and modes, we can understand how certain tonalities have shaped listening trends such as what is popular in 2023. 

The distribution of streams by key shows the preferences by listeners throughout the year. Certain keys such as C major and G major typically dominate the charts, which explains their widespread use in mainstream music. These keys have an open and bright sound, which appeals to both artists and listeners. These popular keys indicate their role in evoking feelings such as energetic uplighting. Another popular key to be streamed is G. Music in this key can create a feeling of peace, calmness, and satisfaction. Top streamed keys can reveal which emotions listeners are looking to evoke. Similar to total streams by musical key, the key C major holds the highest count of songs at 120. Close second are G(count 96) and G major(count 91). The data reveals that certain keys are more commonly used than others, which could be reflective of artists using familiar tonalities that cater towards mainstream appeal. The simplest key to play in piano is C major, making it easy to play. There are also no sharps or flats, contributing to generally bright or cheerful sounds. 

When looking at total streams by musical mode, the distribution of mode provides insight to the overall character of music. Major modes are typically associated with positive, bright emotions. Minor modes are deeper, introspective sounds that may appeal to a smaller range of listeners. In the graph, major has approximately 300,000,000,000 total streams , while minor has close to 200,000,000 total streams. When looking at popular songs for the year, the shift in major vs. minor songs can reveal which emotional state resonated with listeners. 

*Tempo trends: Fast or Slow Beats?*

Factors such as energy, tempo, and danceability play a crucial role in how listeners connect with music. The interplay between these characteristics can transform a song to feel relaxing, groovy, or introspective. A deep dive into how tempo(BPM), danceability, and energy can shape listener experience through the top tracks of 2023.

The tempo distribution shows that most streams fall in the moderate range between approximately 90-120 BPM. This allows for a balance of upbeat and relaxed tempos, suitable for a wide range of moods and genres. There are other spikes around 140,150, and 170 BPM, indicating that there is an appeal for energetic tracks for parties, workouts, and other high energy moments. Slower temps around 80 BPM are typically calming, introspective songs. This wide variation of BPMs showcases that listeners will interact with various tempos depending on activities and emotions. 

The relationship between danceability and energy highlights how songs can balance groove and intensity. Tracks that have moderate energy levels with high danceability often become popular. Listeners are able to engage with the rhythm without overwhelming them. However, songs with high energy and danceability are probably anthems used to cultivate movement and excitement. Songs with lower energy but a high danceability could reflect chill, laid back music like lo-fi or mellow R&B. Conversely, high energy and low danceability could indicate rock or hip-hop. While there is no genre variable for this data, the clusters on this chart may indicate different musical genres for the top songs of 2023.

*The Multidimensional World of Sound*

Music is more than just rhythm and melody - sonic elements help tell stories and create immersive soundscapes. In this section we chose to focus on the balance between liveness, speechiness, and the proportion of vocal and instrumental tracks. Liveness refers to reverberation time, with a live room having a long verbatim time. A higher liveness score indicates there is a higher chance a song was recorded live. These may be live-recordings, acoustic renditions, or unfiltered performances. Speechiness is the prominence of spoken, with a higher value indicating more words spoken in a song. Genres such as rap and spoken-word tracks will likely have higher speechiness values. Most of the values cluster in the low liveness, low speechiness level, mostly studio produced songs. Some songs have a high liveness but have a lower speechiness, likely indicating melodies such as acoustic sessions or crowd-filled songs. No songs have a value for speechiness and liveness, meaning there weren’t any songs that were live with a high percentage of words. 

Both vocal and instrumental songs made their way onto the top tracks of 2023 list. Vocal tracks are those that dominate mainstream music since a listener is able to hear a story and connection through vocal delivery and lyrics. These tracks have the advantage of vocal and instrumental aspects. Pop, R&B, and hip-hop are typical genres that have vocal tracks. Instrumental tracks on the other hand can serve different functions such as relaxation, creating ambient soundscapes, and enhancing concentration. Some of these genres include classical, jazz, and lo-fi, all which lean heavily on instrumentals to develop a sense of mood. This data showed a dominance of vocal tracks at 90.9%. This is expected as vocal tracks are popular in mainstream media, and therefore popular on streaming services such as Spotify. However, 9.13% of the songs reflected on this chart are instrumental, highlighting that there is a large audience for these tracks.

*Data Wrapped: The Ultimate Chart-Toppers of 2023*

In this section, we will take a deep dive into some of the most popular songs for specific variables chosen to be included in this dataset. After all, what is Spotify wrapped without chart-toppers and showstoppers? 

The top ten artists with the most danceable songs included very popular artists from 2023. These tracks are crowd-pleasers used for various occasions such as working out, parties, and jam sessions. Taylor Swift tops the chat with 25 songs, while BTS closed out the chart with 8. These artists reflect a few different styles of music such as pop, Kpop, rap, country, and more. This chart reflects that popularity and danceability may be linked, as the artists listed are some of the most popular artists of the music industry today. These artists helped make the year feel like a continuous dance party with their top hits. The top ten artists with the most songs in 2023 have the exact same group of ten people as the previous chart. Artists who possess the ability to create tracks of both quality and quantity appear on this list. Taylor Swift tops the chart with 34 songs on the list, and BTS ends the chart with 8. The Weekend swoops the number two spot with 22 songs. It is likely that artists that both release more music and have a large level of popularity are more likely to be featured on this chart. 

Next, we looked at the top ten tracks with the most appliances in Spotify charts. Song success can be measured by charts, whether that be global, regional, or genre related. Two songs were from the Barbie movie, a popular film released in the summer of 2023. The top song, Seven, by Jungkook (feat. Latto), is an R&B/Soul, Korea Dance, K-pop song. This variation of different genre types allows for the song to be featured in multiple chart types. Jungkook is part of a popular Korean band BTS and Latto on the other hand is an American rapper and singer. This results in a large range of listeners and regions for the song to grow popular in. Harry Styles has two songs on the list and Taylor Swift has one. Typically songs featured in these charts become cultural touchpoints, relating to widespread success and popularity. 
The top ten most energetic songs can be defined as those with intensity, activity, and overall drive. The most energetic song was a Hip-Hop/Rap track called Murder on my Mind by YNW Melly. The next song was I’m good(blue) which is a dance/electronic, pop song. The chart has various genres included, which can be favored for situations such as festivals, parties, and gyms. The top ten songs in the most playlists takes us out of the year 2023 and highlights some of the most popular songs throughout the years. While these songs weren’t necessarily the top ten most streamed songs of the year, their impact spans across years and even decades. The 5th and 9th songs on the chart, Take on me and Everybody Wants to Rule the World were released in 1985, making them the oldest songs on this list. These songs' continued presence in Spotify playlists speaks volumes about their timeless appeal. Despite competition from newer tracks released in 2023, they remain favorites, highlighting lasting resonance with listeners. It’s a testament to both tracks' quality and relevance, proving they can truly stand the test of time. On the other hand, Blinding Lights was the most recently released song (2020), which has quickly climbed the charts to the 6th most added song to Spotify playlists. This chart gives some insight on which tracks capture a wide appeal, representing some of the most iconic songs of all time. 

#### 2024 Exploratory Analysis

For our exploratory analysis of the 2024 Spotify dataset , we wanted to look more into how track popularity changed across streaming platforms, as well as if the genres of the top songs differed across platforms. The exploratory analysis consisted of four major parts: summary statistics and data (e.g., mean/median/IQR of streams across platforms, percentage of explicit songs, etc.), top 10 artists by different media platforms,  top songs on streaming apps (e.g., Spotify and Soundcloud),  and top songs on socials (e.g., TikTok and YouTube). 

#### Cross-Year Exploratory Analysis

For our exploratory analysis of the merged 2023/2024 Spotify data, we wanted to look more into how track popularity (e.g., TikTok streams) was related to track characteristics (e.g., danceability). The exploratory analysis consisted of four major parts: summary data, summary of musical attributes (e.g., distribution of song count by BPM, distribution of song count by danceability, etc.), musical attributes compared to streaming data (e.g, distribution of Spotify streams by danceability), and musical attributes compared to TikTok data (e.g, distribution of TikTok engagement by danceability).

## Additional Analysis

Using the 2023 and 2024 merged data, we conducted a linear regression assessing the relationship between TikTok engagement and qualitative features of the song such as danceability, energy, and liveness. All variables were treated as numerical for this analysis. The level of significance was set at 5%.  Based on the linear regression assessing the relationship between TikTok engagement and qualitative features of the song such as danceability, energy, and liveness, none of the variables had a statistically significant association with TikTok engagement. However, one interesting insight was that danceability had a positive association with TikTok engagement although it was not statistically significant. This regression was driven by the hypothesis that livelier and energetic music will have more TikTok engagement considering the popularity of TikTok dance challenges. Although the findings were not statistically significant, they were interesting.

## Discussion

#### 2023 Discussion

In order to get a better idea of what listeners gravitated towards on Spotify in 2023, we analyzed the year’s top streamed songs. With variables such as BPM, key, danceability, energy, and more, we crafted a unique listening personality for the year. To bring this analysis to life, we designed a Spotify wrapped inspired page that accurately showcases some of the top hits and highlights that defined 2023. 

When looking at keys and modes, we found that listeners favored C major, G major, G keys, and major over minor mode. D major was found to be the least popular key that year, and the minor mode was streamed 33% less than the major mode. When analyzing tempo trends, we found that listeners most enjoyed the range 90-120 BPM, with other values having slight peaks such as 140, 150, and 170 BPM. Songs that ranked relatively high on the danceability and energy scale were clustered together, indicating a preference towards high intensity anthems. Danceability and energy were compared on a density plot, with the two values lining up almost exactly. This can indicate that danceability and energy are likely pretty similar variables. 

We then looked into the world of sound, analyzing liveness compared to speechiness to uncover if listeners preferred live music and songs with a high percentage of words. We found that most people enjoyed a relatively low value for both, with clusters falling primarily between 0-25  for liveness and 0-60 for speechiness. A pie chart was created to compare instrumental (9.13%)  and vocal (90.9%) songs, with vocal songs being the primary type of top streamed song. 

Top ten charts were created in order to get a better idea of what songs and artists were the most popular across the year for different variables. Artists such as Taylor Swift, Harry Styles, The Weekend, and Drake made multiple appearances across various charts, indicating popularity and large fan bases. The K-Pop band BTS had multiple features and one of their band members created the track featured in the most Spotify charts. When looking at the top ten songs in the most playlist across Spotify for the year, we see that these songs range across various years, allowing insight for what songs have stayed with listeners despite new releases. 

The Spotify Wrapped feature of 2023 introduced 12 unique listening personalities based on users’ Spotify habits. Based on our analysis of the top streamed songs, 2023’s collective musical personality would align with The Luminary. This character is defined by a preference for light, upbeat music. Songs that dominated the charts and playlists were reflective of luminary traits, resonating with listeners who were looking for uplifting, energizing tunes throughout the year!

#### 2024 Discussion

In the 2024 dataset, we found that the mean versus median streams or views across all the platforms are very different from each other, indicating that the distribution of data is skewed for each platform. It is also interesting to note TikTok's popularity, where the platform has both the most mean and median views compared to the other platforms. Spotify has the second greatest amount of mean and median streams for 2024. Additionally, we found that there is almost no overlap among the most popular five songs streamed for Spotify, Soundcloud, YouTube, and TikTok. Our analysis also allowed us to compare the popularity of streaming platforms, where Tik Tok's most viewed song has 233,232,311,463 views, and Soundcloud's most streamed song has 319,835,943 streams. This factor of a hundred difference demonstrates TikTok's popularity compared to other streaming platforms such as SoundCloud. It is also interesting that SoundCloud's top streamed songs are all in the hip hop/rap genre, where Spotify's Top 5 streamed songs generally fall into today's hits/pop.

When looking at the graphs that compare Spotify streams,  YouTube views, Soundcloud streams, and Shazam counts, one can see that increased Tik Tok likes (going from the Low to High category) is associated with increased track streams or views. This makes sense that the more popular songs on streaming platforms or songs that people want to know the name of (through using Shazam) would also be popular on TikTok. 

The top 10 artists with the most streams or views across Spotify, Soundcloud, YouTube, and TikTok vary widely. Streaming platforms like Spotify and TikTok (which have the highest number of user streams) have top artists that more fall under the pop genre, while Soundcloud has almost exclusively rap artists (with the exception of Major Lazer). 

#### Cross-Year Discussion

When analyzing musical attributes (e.g., danceability, liveness, and energy) and Spotify streams (both 2023 and 2024 combined), we found that there appears to be no correlation between increasing Spotify streams and increasing energy and danceability. Songs with both low and high musical attributes (e.g., danceability and energy) have the highest amount of Spotify streams. It appears that songs with lower liveness have less Spotify streams, where the groups of songs with higher liveness generally have higher streams.

When analyzing the TikTok engagement based on the distribution of certain musical attributes, we found that there are some trends in TikTok engagement based on specific attributes. For instance, the highest TikTok engagement was associated with a danceability score of 71, an energy score of 68, and a liveness score of 3. These attributes might be important for TikTok engagement including likes, and views. They might have a predisposition to go viral.

Finally, the linear regression assessing the relationship between TikTok engagement and qualitative features of the song such as danceability, energy, and liveness found that none of the variables have a statistically significant association with TikTok engagement. However, one interesting insight is that danceability has a positive association with TikTok engagement although it is not statistically significant. This might corroborate the associations seen in the danceability plot, which was the highest score with highest TikTok engagement from the plots explored. 

